---
title: Lab Three -- Matrices and Data Frames in R # Title of the document

# Bibliography Details
bibliography: Lab3References.bib  # Name of bib file in the same folder 
bibliographystyle: apa                   # Bibliography style

execute:
  warning: false
  message: false

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab3WriteUp.qmd")

format:
  html: 
    number-sections: TRUE
    code-fold: TRUE
    code-line-numbers: TRUE
    cap-location: top 
  pdf:
    documentclass: article
    number-sections: TRUE
    reference-section-title: "References" 
    code-line-numbers: TRUE
    cap-location: top 
    include-in-header:
      - text: |
          \usepackage{amsmath}
          \usepackage[dvipsnames]{xcolor}
          \usepackage[shortlabels]{enumitem}
          \usepackage{parskip}
          \usepackage{hyperref}
          \usepackage{geometry}
          \usepackage{float}
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
format-links: [pdf]
---
```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})

```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
```

* Complete the tasks below. Make sure to start your solutions in on a new line that starts with "**Solution**:".
* Make sure to use the Quarto Cheatsheet. This will make completing and writing up the lab *much* easier.

Consider the unit square depicted in @fig-unitsquare. 

```{r}
#| label: fig-unitsquare
#| fig-cap: "The unit square."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
p0 <- c(0, 0)
p1 <- c(1, 0)
p2 <- c(1, 1)
p3 <- c(0, 1)
ggplot() +
  geom_segment(aes(x = p0[1], y = p0[2], xend = p1[1], yend = p1[2])) +
  geom_segment(aes(x = p1[1], y = p1[2], xend = p2[1], yend = p2[2])) +
  geom_segment(aes(x = p2[1], y = p2[2], xend = p3[1], yend = p3[2])) +
  geom_segment(aes(x = p3[1], y = p3[2], xend = p0[1], yend = p0[2])) +
  theme_bw() +
  xlim(-0.5, 1.5) +
  ylim(-0.5, 1.5) +
  labs(x = "x", y = "y")
```

The unit square can be defined by two basis vectors
$$\mathbf{v}_1=\begin{bmatrix} 1 \\ 0 \end{bmatrix} ~~~ \textrm{and} ~~~ \mathbf{v}_2=\begin{bmatrix} 0 \\ 1 \end{bmatrix}.$$
From these vectors, we can compute the corner vertices.
\begin{align*}
  p_0 & = (0,0) = (x_0, y_0) \tag{The origin}\\
  p_1 & = v_1 = (x_1, y_1) = (1,0)  \tag{The first basis vector}\\
  p_2 & = v_1 + v_2 = (x_2, y_2)= (1,1) \tag{The sum of basis vectors}\\
  p_3 & = v_2 = (x_3, y_3) =(0,1) \tag{The second basis vector}\\
\end{align*}
Note that the segments that form the boundary go from $p_0$ to $p_1$ to $p_2$ to $p_3$ and back to $p_0$.

Consider the matrix $M$ to be the matrix we get from binding the two basis vectors at the column,
$$\mathbf{M} = \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}.$$

# Question 1
Let's transform $\mathbf{M}$ to demonstrate an interesting linear algebra property.

## Part a
Create the matrix `A` in `R`.
$$\mathbf{A} = \begin{bmatrix} 1 & -7 \\ 5 & 9 \end{bmatrix}$$

**Solution**

Used `matrix()` method to create a matrix `A`
```{r}
#| size: scriptsize
(A<- matrix(c(1,5,-7,9), 2,2))
```


## Part b
Compute the product below in `R` and store it in an object `T`.
$$\mathbf{T} = \mathbf{A}\mathbf{M}.$$

**Solution**

Computed product of the matrix
```{r}
#| size: scriptsize
M<- matrix(c(1,0,0,1), 2,2)
(T<-A%*%M)
```


## Part c
Create `basis.vector.1` (first column of `T`) and `basis.vector.2` (second column of `T`) in `R`.

**Solution**

Created `basis.vector.1` and `basis.vector.2` to store first and second columns
```{r}
#| size: scriptsize
basis.vector.1<-T[,1]
basis.vector.2<-T[,2]
```


## Part d
Compute the points $p_0$, $p_1$, $p_2$, and $p_3$.

**Solution**

Computed the given points
```{r}
#| size: scriptsize
(new_p0=c(0,0))
(new_p1=basis.vector.1)
(new_p2=basis.vector.1+basis.vector.2)
(new_p3=basis.vector.2)
```


## Part e
Copy and paste the code for the unit square. Edit the code to draw the segments that form the boundary based on the points in Part d.

**Solution**

Used `ggplot` to draw the segments that form the boundary based on the points.
```{r}
#| label: fig-unitsquare2
#| fig-cap: "Part 1 with A."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
n.p0 <- new_p0
n.p1 <- new_p1
n.p2 <- new_p2
n.p3 <- new_p3

## used ggplot to draw segments that form the boundary
ggplot() +
  geom_segment(aes(x = n.p0[1], y = n.p0[2], xend = n.p1[1], yend = n.p1[2])) +
  geom_segment(aes(x = n.p1[1], y = n.p1[2], xend = n.p2[1], yend = n.p2[2])) +
  geom_segment(aes(x = n.p2[1], y = n.p2[2], xend = n.p3[1], yend = n.p3[2])) +
  geom_segment(aes(x = n.p3[1], y = n.p3[2], xend = n.p0[1], yend = n.p0[2])) +
  theme_bw() +
  #xlim(-0.5, 1.5) +
  #ylim(-0.5, 1.5) +
  labs(x = "x", y = "y")
```


## Part f
Interestingly, the determinant of $\mathbf{A}$ gives us the area of the shape plotted in Part e. Find the determinant of $\mathbf{A}$.

**Solution**

Used the method `determinant()` to find the determinant of the matrix which also gives us the area.
```{r}
#| size: scriptsize
(determinant<-det(A))
```

# Question 2
Recomplete Question 1 using a new matrix for $\mathbf{A}$. This matrix is not as well behaved.

## Part a
Create the matrix `A` in `R`.
$$\mathbf{A} = \begin{bmatrix} 1 & 2 \\ 2 & 4 \end{bmatrix}$$

**Solution**

Created another matrix A using `matrix()` method

```{r}
#| size: scriptsize
(A<- matrix(c(1,2,2,4), 2,2))
```

## Part b
Compute the product below in `R` and store it in an object `T`.
$$\mathbf{T} = \mathbf{A}\mathbf{M}.$$

**Solution**

Computed its product
```{r}
#| size: scriptsize
(T<-A%*%M)
```

## Part c
Create `basis.vector.1` (first column of `T`) and `basis.vector.2` (second column of `T`) in `R`.

**Solution**

Create `basis.vector.1` (first column of `T`) and `basis.vector.2` (second column of `T`) in `R`.
```{r}
#| size: scriptsize
basis.vector.1<-T[,1]
basis.vector.2<-T[,2]
```


## Part d
Compute the points $p_0$, $p_1$, $p_2$, and $p_3$.

**Solution**

Computed given points
```{r}
#| size: scriptsize
(new_p0=c(0,0))
(new_p1=basis.vector.1)
(new_p2=basis.vector.1+basis.vector.2)
(new_p3=basis.vector.2)
```


## Part e
Copy and paste the code for plotting the unit square. Edit the code to draw the segments that form the boundary based on the points in Part d.

**Note**: As you continue to plot this here and through question 2, you will find that the $x$ and $y$ limits need to change. Try `xlim(-15,15)` and `ylim(-5, 25)`.

**Solution**

Used `ggplot()` to draw the segments that form the boundary based on the points

```{r}
#| label: fig-unitsquare3
#| fig-cap: "Part 2 with new A."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
n.p0 <- new_p0
n.p1 <- new_p1
n.p2 <- new_p2
n.p3 <- new_p3
ggplot() +
  geom_segment(aes(x = n.p0[1], y = n.p0[2], xend = n.p1[1], yend = n.p1[2])) +
  geom_segment(aes(x = n.p1[1], y = n.p1[2], xend = n.p2[1], yend = n.p2[2])) +
  geom_segment(aes(x = n.p2[1], y = n.p2[2], xend = n.p3[1], yend = n.p3[2])) +
  geom_segment(aes(x = n.p3[1], y = n.p3[2], xend = n.p0[1], yend = n.p0[2])) +
  theme_bw() +
  #xlim(-15, 15) +
  #ylim(-5, 25) +
  labs(x = "x", y = "y")
```


## Part f
Interestingly, the determinant of $\mathbf{A}$ gives us the area of the shape plotted in Part e. Find the determinant of $\mathbf{A}$.

**Solution**

Used `determinant()` method to find the deteminant
```{r}
#| size: scriptsize
(determinant<-det(A))
```


# Question 3
We can conduct a shear transformation that keeps the area of the shape the same but horizontally or vertically slants the object. 

In fact, this is one of the ways artists can make something look like it's leaning or being pushed. Other applications where this is useful include fluid dynamics and crystallography, where the area of a deformed object must remain constant, or in photography where perspective in photos can be adjusted.

A shear transformation is implemented by altering the basis vectors we start with. For vertical slanting,
$$\mathbf{M}_v = \begin{bmatrix} 1 & k \\ 0 & 1 \end{bmatrix}$$
and for horizontal slanting
$$\mathbf{M}_h = \begin{bmatrix} 1 & 0 \\ k & 1 \end{bmatrix}.$$
In both cases, $k \in \mathbb{R}$ is called a shear factor and its sign determines the direction of the slant and its magnitude determines the severity of the slant.

## Part a
Use the following to generate a new shape. Compare the shape and its area to your answer in Question 1.
$$\mathbf{M}_v = \begin{bmatrix} 1 & k=0.5 \\ 0 & 1 \end{bmatrix} ~~~~~ \textrm{and} ~~~~~
  \mathbf{A}   = \begin{bmatrix} 1 & -7 \\ 5 & 9 \end{bmatrix}$$

**Solution**

Generated a vertically cheared shape
```{r}
#| size: scriptsize
M<- matrix(c(1,0,0.5,1), 2,2)
A<- matrix(c(1,5,-7,9), 2,2)
(T<-A%*%M)
basis.vector.1<-T[,1]
basis.vector.2<-T[,2]
(new_p0=c(0,0))
(new_p1=basis.vector.1)
(new_p2=basis.vector.1+basis.vector.2)
(new_p3=basis.vector.2)
```

Area
```{r}
#| size: scriptsize
(determinant<-det(T))
```

Shape
```{r}
#| label: fig-unitsquare4
#| fig-cap: "Vertical shearing."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
n.p0 <- new_p0
n.p1 <- new_p1
n.p2 <- new_p2
n.p3 <- new_p3
ggplot() +
  geom_segment(aes(x = n.p0[1], y = n.p0[2], xend = n.p1[1], yend = n.p1[2])) +
  geom_segment(aes(x = n.p1[1], y = n.p1[2], xend = n.p2[1], yend = n.p2[2])) +
  geom_segment(aes(x = n.p2[1], y = n.p2[2], xend = n.p3[1], yend = n.p3[2])) +
  geom_segment(aes(x = n.p3[1], y = n.p3[2], xend = n.p0[1], yend = n.p0[2])) +
  theme_bw() +
  #xlim(-15, 15) +
  #ylim(-5, 25) +
  labs(x = "x", y = "y")
```

## Part b
Use the following to generate a new shape. Compare the shape and its area to your answer in Questions 1 and the vertical shear transform in part (a).
$$ \mathbf{M}_h = \begin{bmatrix} 1 & 0 \\ k=0.5 & 1 \end{bmatrix}~~~~~ \textrm{and} ~~~~~
  \mathbf{A}   = \begin{bmatrix} 1 & -7 \\ 5 & 9 \end{bmatrix}$$

**Solution**

Created a horizontally sheared shape

```{r}
#| size: scriptsize
M<- matrix(c(1,0.5,0, 1), 2,2)
A<- matrix(c(1,5,-7,9), 2,2)
(T<-A%*%M)
basis.vector.1<-T[,1]
basis.vector.2<-T[,2]
(new_p0=c(0,0))
(new_p1=basis.vector.1)
(new_p2=basis.vector.1+basis.vector.2)
(new_p3=basis.vector.2)
```

Area
```{r}
#| size: scriptsize
(determinant<-det(T))
```

Shape
```{r}
#| label: fig-unitsquare5
#| fig-cap: "Horizontal shearing."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
n.p0 <- new_p0
n.p1 <- new_p1
n.p2 <- new_p2
n.p3 <- new_p3
ggplot() +
  geom_segment(aes(x = n.p0[1], y = n.p0[2], xend = n.p1[1], yend = n.p1[2])) +
  geom_segment(aes(x = n.p1[1], y = n.p1[2], xend = n.p2[1], yend = n.p2[2])) +
  geom_segment(aes(x = n.p2[1], y = n.p2[2], xend = n.p3[1], yend = n.p3[2])) +
  geom_segment(aes(x = n.p3[1], y = n.p3[2], xend = n.p0[1], yend = n.p0[2])) +
  theme_bw() +
  #xlim(-15, 15) +
  #ylim(-5, 25) +
  labs(x = "x", y = "y")
```

## Part c
Can we shear vertically and horizontally at the same time? Use the following to generate a new shape. Compare the shape and its area to your answers in parts (a) and (b).
$$\mathbf{M}_{vh} = \begin{bmatrix} 1 & k=0.5 \\ k=0.5 & 1 \end{bmatrix} ~~~~~ \textrm{and} ~~~~~
  \mathbf{A}   = \begin{bmatrix} 1 & -7 \\ 5 & 9 \end{bmatrix}$$

**Solution**

Created a vertically and horizontally sheared shape
```{r}
#| size: scriptsize
M<- matrix(c(1,0.5,0.5, 1), 2,2)
A<- matrix(c(1,5,-7,9), 2,2)
(T<-A%*%M)
basis.vector.1<-T[,1]
basis.vector.2<-T[,2]
(new_p0=c(0,0))
(new_p1=basis.vector.1)
(new_p2=basis.vector.1+basis.vector.2)
(new_p3=basis.vector.2)
```

Area
```{r}
#| size: scriptsize
(determinant<-det(T))
```

Shape
```{r}
#| label: fig-unitsquare6
#| fig-cap: "Vertical and horizontal shearing."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
n.p0 <- new_p0
n.p1 <- new_p1
n.p2 <- new_p2
n.p3 <- new_p3
ggplot() +
  geom_segment(aes(x = n.p0[1], y = n.p0[2], xend = n.p1[1], yend = n.p1[2])) +
  geom_segment(aes(x = n.p1[1], y = n.p1[2], xend = n.p2[1], yend = n.p2[2])) +
  geom_segment(aes(x = n.p2[1], y = n.p2[2], xend = n.p3[1], yend = n.p3[2])) +
  geom_segment(aes(x = n.p3[1], y = n.p3[2], xend = n.p0[1], yend = n.p0[2])) +
  theme_bw() +
  #xlim(-15, 15) +
  #ylim(-5, 25) +
  labs(x = "x", y = "y")
```

# Question 4

## Part a
Create a data frame `ladder` with the following data. Note your data frame should have three columns; I split it to save vertical space here.

| x    | y    | group| x    | y    | group|
|:-----|:-----|:-----|:-----|:-----|:-----|
| 0    | 0    |    1 | 1    | 8    |    6 |
| 0    | 20   |    1 | 0    | 10   |    7 |
| 1    | 0    |    2 | 1    | 10   |    7 |
| 1    | 20   |    2 | 0    | 12   |    8 |
| 0    | 2    |    3 | 1    | 12   |    8 |
| 1    | 2    |    3 | 0    | 14   |    9 |
| 0    | 4    |    4 | 1    | 14   |    9 |
| 1    | 4    |    4 | 0    | 16   |   10 |
| 0    | 6    |    5 | 1    | 16   |   10 |
| 1    | 6    |    5 | 0    | 18   |   11 |
| 0    | 8    |    6 | 1    | 18   |   11 |


**Solution**

Created dataframe `ladder` with the above specifications.

```{r}
#| size: scriptsize
(group.dat<-rep(x=1:11, each=2))
(x.dat<-c(c(0,0,1,1),rep(c(0,1), times=9)))
(y.dat<-c(c(0,20,0,20),rep(seq(2,18, 2), each=2)))
(ladder<-data.frame(x=x.dat, y=y.dat, group=group.dat))
```

## Part b
Remove `#|eval: false` in the code below. The result should look like a ladder and a very rectangular house if part a is correct.

```{r}
#| eval: false
#| label: fig-ladder
#| fig-cap: "A ladder."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
ggplot() +
  geom_path(data = ladder, 
              aes(x = x, y = y, group = group)) +
                  geom_segment(aes(x = 10, xend = 10,
                   y=0, yend = 20))+
  geom_segment(aes(x = 10, xend = 20,
                   y=20, yend = 20))+
  geom_segment(aes(x = 20, xend = 20,
                   y=0, yend = 20))+
  geom_segment(aes(x = 10, xend = 20,
                   y=0, yend = 0)) +
  theme_bw() +
  xlim(-2,22) +
  ylim(-2,22)
```

**Solution**

Removed `eval: false` from the code to display the ladder and rectangular house

```{r}
#| label: fig-ladder1
#| fig-cap: "A ladder."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
ggplot() +
  geom_path(data = ladder, 
              aes(x = x, y = y, group = group)) +
                  geom_segment(aes(x = 10, xend = 10,
                   y=0, yend = 20))+
  geom_segment(aes(x = 10, xend = 20,
                   y=20, yend = 20))+
  geom_segment(aes(x = 20, xend = 20,
                   y=0, yend = 20))+
  geom_segment(aes(x = 10, xend = 20,
                   y=0, yend = 0)) +
  theme_bw() +
  xlim(-2,22) +
  ylim(-2,22)
```

## Part c
Use `as.matrix()` to create a $22 \times 2$ matrix $\mathbf{A}$ containing the `x` and `y` observations from the `ladder` data frame. Use matrix multiplication to compute
$$\mathbf{T} = \mathbf{A} \mathbf{M}_h.$$
Recall, 
$$\mathbf{M}_h = \begin{bmatrix} 1 & 0 \\ k & 1 \end{bmatrix}$$
and use $k=0.5.$

**Solution**

Created a $22 \times 2$ matrix and found the product.

```{r}
#| size: scriptsize
(A<- as.matrix(ladder[,1:2]))
M<-matrix(c(1,0.5,0.1),2,2)
(T=A%*%M) # product
```

## Part d
Create a data frame `leaning.ladder` with the `x` and `y` equal to the first and second column of $\mathbf{T}$, respectively, and the same values of `group` from `ladder`. 

**Solution**

Created a new dataframe with `x` and `y` equal to first and second column of T.

```{r}
#| size: scriptsize
(leaning.ladder<-data.frame(x=T[,1], y=T[,2], group=group.dat))
```

## Part e
Remove `eval: false` in the code below. The result should look like a ladder and a very rectangular house if part a is correct. The new leaning ladder should look like it is leaning on the house.

```{r}
#| eval: false
#| label: fig-ladder2
#| fig-cap: "A ladder and a leaning ladder."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
ggplot() +
  geom_path(data = ladder, 
              aes(x = x, y = y, group = group, color = "Starting Ladder")) +
  geom_path(data = leaning.ladder, 
              aes(x = x, y = y, group = group, color = "Leaning Ladder")) +
  geom_segment(aes(x = 10, xend = 10,
                   y=0, yend = 20))+
  geom_segment(aes(x = 10, xend = 20,
                   y=20, yend = 20))+
  geom_segment(aes(x = 20, xend = 20,
                   y=0, yend = 20))+
  geom_segment(aes(x = 10, xend = 20,
                   y=0, yend = 0)) +
  theme_bw() +
  xlim(-2,22) +
  ylim(-2,22) +
  scale_color_manual("", values=c("darkred", "grey"))
```

# Solution:

Removed `eval: false` from code to display figures.
```{r}
#| label: fig-ladder3
#| fig-cap: "A ladder and a leaning ladder."
#| fig-width: 3
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: "scriptsize"
ggplot() +
  geom_path(data = ladder, 
              aes(x = x, y = y, group = group, color = "Starting Ladder")) +
  geom_path(data = leaning.ladder, 
              aes(x = x, y = y, group = group, color = "Leaning Ladder")) +
  geom_segment(aes(x = 10, xend = 10,
                   y=0, yend = 20))+
  geom_segment(aes(x = 10, xend = 20,
                   y=20, yend = 20))+
  geom_segment(aes(x = 20, xend = 20,
                   y=0, yend = 20))+
  geom_segment(aes(x = 10, xend = 20,
                   y=0, yend = 0)) +
  theme_bw() +
  xlim(-2,22) +
  ylim(-2,22) +
  scale_color_manual("", values=c("darkred", "grey"))
```