---
title: Lab Eight -- Probability # Title of the document

# Bibliography Details
bibliography: Lab8References.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab8WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---
```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})
```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(patchwork)
library(extraDistr)
```

**Tips**

* Complete the tasks below. Make sure to start your solutions in on a new line that starts with "**Solution**:".
* Make sure to use the Quarto Cheatsheet. This will make completing and writing up the lab *much* easier.
* Make sure to use "enter" to make your code readable, so it doesn't run off the page. I switched the Quarto document to automatically render to .pdf, so you can see what you're submitting by default. You can switch back for the highlighting by moving the html block above the pdf block in the YAML header.
* Don't forget that you can copy-and-paste code when you're doing something similar as we did in class!
* Make sure to plot all computed probabilities. This is good `ggplot` practice AND will help make sure you don't make a mistake when computing the probability.

# The Binomial and Negative Binomial Distributions

## The Binomial Distribution
In many introductory courses, we make the simplifying assumption of independence in Bernoulli trials. For example, when playing best two-out-of-three legs in a darts match, we would assume Player A has an 70% chance of winning each leg. We write:
$$X \sim \text{Binomial}(n=3, p=0.70)$$
where

* $X$ is the number of legs won
* $n$ is the number of trials 
* $p$ is the probability of the event of interest

Of note, it is not necessarily the case that the players play all three legs -- if either player wins legs one and two the match ends.

There are eight possible outcomes for Player A:

* (win, win, "win") $\longrightarrow$ Player A wins
* (win, win, "lose") $\longrightarrow$ Player A wins
* (win, lose, win) $\longrightarrow$ Player A wins
* (win, lose, lose) $\longrightarrow$ Player A loses
* (lose, win, win)  $\longrightarrow$ Player A wins
* (lose, win, lose) $\longrightarrow$ Player A loses
* (lose, lose, "win") $\longrightarrow$ Player A loses
* (lose, lose, "lose") $\longrightarrow$ Player A loses

**Note**: I use quotations above to denote outcomes where the last match isn't actually played.

**Use the Binomial distribution to compute the probability that Player A wins the match.  Make sure write out your probability statement and work to write it in terms of the distribution function (PMF/CDF) to plot the probability using `ggplot`.**

**Solution:**

Used `dbinom` to find the binomial probabilty distribution in which player A wins the three games and plotted both the CDF and PMF with `ggplot(...)`. Calculated the probability that A wins using the CDF `P(X>=2)=1-P(X<2)=1-P(X<=1)` and plotted the PMF and CDF.
```{r}
#|size: scriptsize

# tibble with PMF & PDF valuesfor each x from 0:3
gg.dat<-tibble(x= 0:3) |>
  mutate(PMF= dbinom(x, size=3, prob = 0.70),
         CDF= pbinom(x, size=3, prob=0.70))

# probability A wins- CDF
p.win <- sum(dbinom(2:3, size = 3, prob = 0.70))
(p.win)

# PMF plot
PMF.plot<-ggplot(data=gg.dat)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probabiity that A wins")

# CDF plot
CDF.plot<- ggplot(data=gg.dat) +
  geom_step(aes(x=x, y=CDF))+
  geom_point(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability")

PMF.plot | CDF.plot
```

## The Negative Binomial Distribution
In intermediate courses, we learn the Negative Binomial distribution that we use to model the number of "failures" until the $r$th success. Note that this approach still makes the simplifying assumption of independence in Bernoulli trials.

**Specify the two cases necessary to compute the probability. Fully explain these probability statements in the context of a best two-out-of-three match and the Negative Binomial distribution. Finally, compute the probability that Player A wins the match and show you get the same result as the Binomial approach. Make sure write out your probability statement and work to write it in terms of the distribution function (PMF/CDF) to plot the probability using `ggplot`.**

**Solution:**
Found the number of times that player A loses the game before actually winning the match. Number of successes in this game is `(r=2)`. Finding the negative binom is `X=negativeBinom(r=2, p=0.70)`

```{r}
#|size: scriptsize
gg.dat<-tibble(x= -1:10) |>
  mutate(PMF= dnbinom(x, size=3, prob = 0.70),
         CDF= pnbinom(x, size=3, prob=0.70))


PMF.plot<-ggplot(data=gg.dat)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probabiity")+
  xlab("Number of failures before 2nd win")

CDF.plot<- ggplot(data=gg.dat) +
  geom_step(aes(x=x, y=CDF))+
  geom_point(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Number of failure before 2nd win")

PMF.plot | CDF.plot

```

**Calculation**
`P(X=0)+P(X=1)`:

```{r}
# size: scriptsize
# P(X=0)
p0 <- dnbinom(0, size = 2, prob = 0.70)

# p(X=1)
p1 <- dnbinom(1, size = 2, prob = 0.70)

#probability A wins
p_win <- p0 + p1
p_win
```

Binom approach=negative Binom approach= 0.784

# The Beta Binomial Distribution
One approach to generalize this modeling is to use a Beta-Binomial distribution. This distribution arrises when we let $p$ be a random variable that follows a Beta distribution:
\begin{align*}
  P &\sim \text{Beta}(\alpha, \beta)\\
  X &\sim \text{Binomial}(n, P).
\end{align*}
This distribution is not part of base `R`, but is available using the `_bbinom` functions from the `extraDistr` package [@extraDistr].

The primary benefit is that the Beta-Binomial enables us to specify some uncertainty about the skill of Player A, as $p$ does not have to be fixed (e.g., $p=0.70$). Suppose the result of the best two-out-of-three match follows a Beta-Binomial distribution  where $\alpha=7$ and $\beta=3$.

## The Beta Part
1. **Plot the corresponding Beta distribution. Interpret the plot in the context of a best two-out-of-three match and the Binomial distribution.**\
**Hint**: The Beta distribution is available in base `R` via `_beta()`.

**Solution:**

Plotted the PDF and CDF of the corresponding beta distribution. In the context of a best two-out-of-three match, it means the probability that Player A wins each leg is not assumed to be exactly 0.70 every time. The Beta distribution represents our uncertainty about Player A’s ability.

```{r}
#|size: scriptsize

# aplha & beta values
alpha=7
beta=3

# continous data on PDF & CDF
gg.dat<-tibble(x= seq(0, 1, length.out=1000)) |>
  mutate(PDF= dbeta(x, alpha, beta),
         CDF= pbeta(x, alpha, beta))

# PDF plot
PDF.plot<-ggplot(data=gg.dat)+
  geom_line(aes(x=x,y=PDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Density")
  
# CDF plot
CDF.plot<- ggplot(data=gg.dat) +
  geom_line(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability")

PDF.plot | CDF.plot
```

1. How does the Beta incorporate uncertainty about the skill of Player A? **Compute the mean and variance of the distribution.**\
**Hint**: You will find the `integrate(...)` function helpful here. 
\begin{align*}
  E(P) &= \mu_P = \int_{0}^1 p f_P(p) dp\\
  var(P) &= \sigma^2_P = \int_{0}^1 (p-\mu_P)^2 f_P(p) dp\\
\end{align*}

**Solution:**

Created functions to compute mean and variance and used the `integrate(...)` function to integrate them from 0 to 1.

```{r}
#|size: scriptsize

# funciton to calculate mean
mean.val <- function(p) {
  p * dbeta(p, alpha, beta)
  }
# integrate from 0 to 1
(mu <- integrate(mean.val, lower = 0, upper = 1)$value)
# funciton to calculate variance
variance.val<-function(p){
  (p-mu)^2 * dbeta(p, alpha, beta)
}
# integrate from 0 to 1
(variance <- integrate(variance.val, lower = 0, upper = 1)$value)
```

1. **What is the probability that Player A has a win probability of exactly 0.70?  Make sure write out your probability statement and work to write it in terms of the distribution function (PDF/CDF) to plot the probability using `ggplot`.**\
**Hint:** Think carefully about the properties of continuous distributions here.

**Solution:**

The probability of player A having a win probability of 0.70 is `P(X=0.70)= 0` because area under the PDF curve for a single point is 0 since it has no width.

```{r}
#|size: scriptsize

# continous (x) for PDF and CMF
gg.dat<-tibble(x= seq(0, 1, length.out=1000)) |>
  mutate(PDF= dbeta(x, alpha, beta),
         CDF= pbeta(x, alpha, beta))

# PDF plot
PDF.plot<-ggplot(data=gg.dat)+
  geom_line(aes(x=x,y=PDF))+
  geom_linerange(data=tibble(x=0.70)|> mutate(PDF=dbeta(x,alpha,beta)),
  aes(x= x,ymin=0, ymax=PDF),color="darkorchid2")+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Density")

# CDF plot
CDF.plot<- ggplot(data=gg.dat) +
  geom_line(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability")

PDF.plot | CDF.plot
```

1. **What is the probability that Player A has a win probability of at least 0.50?  Make sure write out your probability statement and work to write it in terms of the distribution function (PDF/CDF) to plot the probability using `ggplot`.**

**Solution:**

The probability of player A having a win probability of at least 0.50 is `P(X>=0.50) = 1-P(X<0.50)= 1-P(X<=0.50)` which is same as `F(x)` in the CDF

```{r}
#|size: scriptsize

# contious x for PDF and CDF in beta distrib
gg.dat<-tibble(x= seq(0, 1, length.out=1000)) |>
  mutate(PDF= dbeta(x, alpha, beta),
         CDF= pbeta(x, alpha, beta))

# probability of A winning (>=0.5)
p.win=1 - pbeta(0.50, 7, 3)
(p.win)

# PDF plot
PDF.plot<-ggplot(data=gg.dat)+
  geom_line(aes(x=x,y=PDF))+
  geom_ribbon(data= gg.dat|> filter(x> 0.5),aes(x=x, ymin=0, ymax=PDF),fill="darkorchid2", alpha=0.50)+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Density")

# CDF plot
CDF.plot<- ggplot(data=gg.dat) +
  geom_line(aes(x=x, y=CDF))+
  geom_point(data=tibble(x=0.5) |>mutate(CDF=pbeta(x,alpha,beta)), aes(x=x,y=CDF),color="darkorchid2")+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability")

PDF.plot | CDF.plot
```

1. **What is the 90th percentile of Player A's win probability based on this Beta distribution? Make sure write out your probability statement and work to write it in terms of the distribution function (PDF/CDF) to plot the percentile using `ggplot`.**


**Solution:**

The 90th percentile of player A's win probability is `P(X=x)=0.90` and is found by using the inverse CDF function `qbeta(..)` and plootted the distribution by inversing the CDF plot

```{r}
#| size: scriptsize

# created a tibble with two columns, PMF and CDF
gg.dat <- tibble(x = seq(0, 1, length.out = 1000)) |>
  mutate(PDF = dbeta(x, alpha, beta),
        CDF = pbeta(x, alpha, beta))

# inverse using qbeta(...)
inv<-qbeta(0.90, alpha, beta)
(inv)

# Plotted the inverse cdf to show value in 90th percentile
inv.CDF.plot <- ggplot(data = gg.dat) +
  geom_line(aes(x = CDF, y = x)) +
  geom_point(data = tibble(CDF = 0.90) |> mutate(x = qbeta(CDF, alpha, beta)), aes(x = CDF, y = x), color = "darkorchid2") +
  theme_bw() +
  xlab("Cumulative Probability") +
  ylab("x")

inv.CDF.plot
```


## The Beta-Binomial Part
**Compute the probability that Player A wins the match. Compare your result to the Binomial and Negative Binomial results. Make sure write out your probability statement and work to write it in terms of the distribution function (PMF/CDF) to plot the probability using `ggplot`**

**Solution:**

Finding `X-Beta-Binomial(n=3,alpha=7,beta=3)`. `P(X>=2)=1-P(X<2)=1-P(X<=1)` then finding the CDF and `P(X<=1)` and subtracting from 1. 

```{r}
#|size: scriptsize

# created a tibble with two columns, PMF and CDF
gg.dat<-tibble(x= -1:10) |>
  mutate(PMF= dbbinom(x, size=3, alpha, beta),
         CDF= pbbinom(x, size=3, alpha, beta))

# 1-P(X<=1)
p.win <- 1 - pbbinom(1, size=3, alpha, beta)
(p.win)

# PMF plot using linerange
PMF.plot<-ggplot(data=gg.dat)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probability that A wins")

# CDF plot shown in steps
CDF.plot<- ggplot(data=gg.dat) +
  geom_step(aes(x=x, y=CDF))+
  geom_point(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability")

PMF.plot | CDF.plot
```

## The Why
Check the expected value and variance of each distribution.

Analytically, we have
\begin{align*}
  E(X) &= np  &&var(X) = np(1-p) \tag{Binomial Distribution}\\
  E(X) &= n\left(\frac{\alpha}{\alpha+\beta}\right)  &&var(X) = \frac{n \alpha \beta (\alpha + \beta + n)}{(\alpha+\beta)^2(\alpha + \beta + 1)} \tag{Beta-Binomial Distribution}
\end{align*}

**For each distribution, compute the mean number of legs won for player A.**\
**Hint**: You can perform the sum over the finite support set here. 
\begin{align*}
  E(X)   = \mu_X    &= \sum_{i=0}^3 i f_X(i)\\
  var(X) = \sigma_X &= \sum_{i=0}^3 (i-\mu_X) f_X(i)
\end{align*}

**Solution:**

Created functions to compute the mean and variance by using the formula and summed using the `sum()` function the applied values 0 to 3 using `sapply(...)`
```{r}
#|size: scriptsize

# function to mean calculate given the formula
mean.val <- function(i) {
  i * dbbinom(i, 3, alpha, beta) # i * beta binomial
  }
(mu <- sum(sapply(0:3, mean.val))) ## use sapply to find sum to 3
# function to calculate variance given the formula
variance.val<-function(i){
  (i-mu)^2 * dbbinom(i, 3, alpha, beta)
}
(variance <- sum(sapply(0:3, variance.val)))  ## use sapply to find sum to 3
```

# Altham's Multiplicative Binomial Distribution
While the Beta-Binomial enables us to model cases where winning has a clustering effect (e.g., a "hot hand").

## Probability Mass Function
The mass function for this distribution is below. While it used to be available in the `MBD` package for `R`, this package is no longer available.
$$f_X(x) = \frac{1}{c} {n \choose x} p^x (1-p)^{n-x} \theta^{x(n-x)}I(x \in \{0, 1, ..., n\})$$
where
$$c = \sum_{i=0}^n {n \choose i} p^i (1-p)^{n-i} \theta^{i(n-i)}$$ 

**Write a function in `R` called `daltbinom()` for computing the PMF of this distribution. Use your function to plot the distribution where $p=0.70$ where $\theta=0.50$, $\theta=1$, and $\theta=1.50$. Comment on the differences in the context of a best two-out-of-three match**


**Solution:**

Creating a function `daltbinom()` with arguments `x`, `size` for number of observations, `p` for probability, and `t` for theta then implemented the given MF formula to obtain the PMF. Used that function to plot the PMF of the data for different theta values.

```{r}
#|size: scriptsize

# daltbinom function for computing the PMF given the formula and argument values
daltbinom<-function(x, size, p, t){
    i=0:size # for summation from vals 0 to size
    res = sum(p^(i) * (1-p)^(size-(i)) * t^((i)*(size-(i)))) # summation for the formula (c)
    ifelse(x >= 0 & x<= size, ##condition
    (p^x * (1-p)^(size-x) * t^(x*(size-x))) / res, 0) # only calculate (formula/c) if x is between 0(inclusive) and size(inclusive)
}

# created variables with PMF values with different theta values
gg.dat1<-tibble(x= -1:10) |>
  mutate(PMF= daltbinom(x, size=3, 0.70, 0.5))

gg.dat2<-tibble(x= -1:10) |>
  mutate(PMF= daltbinom(x, size=3, 0.70, 1))

gg.dat3<-tibble(x= -1:10) |>
  mutate(PMF= daltbinom(x, size=3, 0.70, 1.5))

# PMF plot for when theta is 0.5
PMF.plot1<-ggplot(data=gg.dat1)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probability that A wins for theta=0.5")

# PMF plot for when theta is 1
PMF.plot2<-ggplot(data=gg.dat2)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probability that A wins for theta=1")

# PMF plot for when theta is 1.5
PMF.plot3<-ggplot(data=gg.dat3)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probability that A wins for theta=1.5")

PMF.plot1 | PMF.plot2 | PMF.plot3
```


## Cumulative Distribution Function
**Use your `daltbinom()` function to create a `paltbinom()` function for computing the CDF of this distribution. Use your function to plot the distribution where $p=0.70$ where $\theta=0.50$, $\theta=1$, and $\theta=1.50$.**

**Solution:**

Created a `paltbinom()` function to compute the CDF using the PMF values obtained from the function `daltbinom()`. Summed all the values from 0 to i to obtain this and used vectorized `ifelse()` function to mormalise to 0 if > 0 or 1 if >=size

```{r}
#|size: scriptsize

# summation helper function to calculate the sum(CDF)
summation<-function(i, size, p, t){ 
  vals=0:i # from 0 to i
  sum(daltbinom(vals, size, p, t)) #summing the pdf from 0 to i
}

#paltbinom function for CDF
paltbinom <- function(x, size, p, t){
  ifelse(x < 0, 0, ifelse(x >= size, 1, sapply(x, summation, size, p, t)))
 # 0 if x is less than 0, else if greater than size, 1. Apply daltbinom if not less than 0 or greater than/equal to size
}

# created variables with PMF values with different theta values
gg.dat1<-tibble(x= -1:10) |>
  mutate(CDF= paltbinom(x, size=3, p=0.70, t=0.5))

gg.dat2<-tibble(x= -1:10) |>
  mutate(CDF= paltbinom(x, size=3, p=0.70, t=1))

gg.dat3<-tibble(x= -1:10) |>
  mutate(CDF= paltbinom(x, size=3, p=0.70, t=1.5))

# CDF plot for when theta is 0.5
CDF.plot1<- ggplot(data=gg.dat1) +
  geom_step(aes(x=x, y=CDF))+
  geom_point(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability for theta=0.5")

# CDF plot for when theta is 1
CDF.plot2<- ggplot(data=gg.dat2) +
  geom_step(aes(x=x, y=CDF))+
  geom_point(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability for theta=1.0")

# CDF plot for when theta is 1.5
CDF.plot3<- ggplot(data=gg.dat3) +
  geom_step(aes(x=x, y=CDF))+
  geom_point(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability for theta=1.5")

CDF.plot1 | CDF.plot2 | CDF.plot3
```

## The Winning Probability
**Compute the probability that Player A wins the match using Altham's Multiplicative Binomial Distribution. Compare your result to the Binomial and Negative Binomial results. Compare your results to the Beta-Binomial model, too. Make sure write out your probability statement and work to write it in terms of the distribution function (PMF/CDF) to plot the probability using `ggplot`.**

**Solution:**

Finding `P(X>=2)=P(X<2)=1−P(X<=1)`. Used the CDF at `P(x<=1)` and subtracted it from one to find the winning probability then plotted the PMF and CDF  

```{r}
#|size: scriptsize

# created a tibble with two columns, PMF and CDF
gg.data<-tibble(x= -1:10) |>
  mutate(PMF= daltbinom(x, size=3, 0.70, 0.5),
         CDF= paltbinom(x, size=3, 0.70, 0.5))

# 1−P(X<=1)
p.wins<-1 - paltbinom(1, size = 3, p = 0.70, t = 0.5)
(p.wins)

# PMF plot using linerange
PMF.plot<-ggplot(data=gg.data)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probability that A wins")

# CDF plot shown in steps
CDF.plot<- ggplot(data=gg.data) +
  geom_step(aes(x=x, y=CDF))+
  geom_point(aes(x=x, y=CDF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Cumulative Probability")

PMF.plot | CDF.plot
```


# Altham's Multiplicative Beta-Binomial Distribution
You may have wondered -- "Can we do the Beta thing with Altham's version of the Binomial Distribution?" The answer is yes!

## Probability Mass Function
The mass function for this distribution is below. 
$$f_X(x) = \left[\int_{0}^1 dp
\left(\frac{1}{c} {n \choose x} p^x (1-p)^{n-x} \theta^{x(n-x)}\right)\left(\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)} p^{\alpha-1}(1-p)^{\beta-1}\right)\right]I(x \in \{0, 1, ..., n\})$$

**Write a function in `R` called `daltbbinom()` for computing the PMF of this distribution. Use your function to plot the distribution where $\alpha=7$ and $\beta=3$ where $\theta=0.50$, $\theta=1$, and $\theta=1.50$. Comment on the differences in the context of a best two-out-of-three match.**\

**Hint**: You can use the `integrate()` in `R` to compute the integral. Note that this requires the function to be vectorized -- it should be unless you have added any block conditionals. 

**Solution:**

Created a new `daltbbinom()` function to compute the Beta-mixed PMF. Generated plots for different theta values and used `sapply()` to evaluate the function across x values.

```{r}
#|size: scriptsize

# daltbbinom function for computing the PMF given the formula and argument values
daltbbinom <- function(x, size, t, a, b){
  if (x < 0 || x > size) return(0)
  integrand <- function(p){ ## create function to calculate given the formula
    daltbinom(x, size, p, t) * dbeta(p, a, b) # daltbinom * dbeta- formula
  }
  integrate(integrand, lower = 0, upper = 1)$value # integrate from 0 to 1
}

#given alpha and beta values
alph=7
bet=3

# created variables with PMF values with different theta values using sapply to apply the x values in the function
gg.dat1 <- tibble(x = -1:10) |>
  mutate(PMF = sapply(x, daltbbinom, size = 3, t = 0.5, a = alph, b = bet)) 
gg.dat2 <- tibble(x = -1:10) |>
  mutate(PMF = sapply(x, daltbbinom, size = 3, t = 1, a = alph, b = bet)) 
gg.dat3 <- tibble(x = -1:10) |>
  mutate(PMF = sapply(x, daltbbinom, size = 3, t = 1.5, a = alph, b = bet))

# PMF plot for when theta is 0.5
PMF.plot1<-ggplot(data=gg.dat1)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probability that A wins for theta=0.5")

# PMF plot for when theta is 1
PMF.plot2<-ggplot(data=gg.dat2)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probability that A wins for theta=1")

# PMF plot for when theta is 1.5
PMF.plot3<-ggplot(data=gg.dat3)+
  geom_linerange(aes(x=x, ymin=0, ymax=PMF))+
  geom_hline(yintercept=0)+
  theme_bw()+
  ylab("Probability that A wins for theta=1.5")

PMF.plot1 | PMF.plot2 | PMF.plot3
```

# A Comparison
Compute the probabilities for all outcomes for Player A.

* Loses 0-2 (LL"L", LL"W")
* Loses 1-2 (WLL, LWL)
* Wins 2-0 (WW"W", WW"L")
* Wins 2-1 (LWW, WLW)

**Note**: You'll need to define these outcomes in terms of the random variable. 

**Create a bar plot (or column plot) using `ggplot` to show the probability of each outcome under the following conditions:**

* **Binomial**
* **Negative Binomial**\
**Note**: You can compute the lose probabilities the same way as the win probabilities using success probability ($1-p$)
* **Beta-Binomial**
* **Althman Binomial with $\theta=0.50$ and $\theta=1.5$**
* **Althman Beta-Binomial with $\theta=0.50$ and $\theta=1.5$**

**Comment on the differences in the context of a best two-out-of-three match.**

**Solution**

Computed probabilities for all listed outcomes using all the distriibutions and stored them in vectors then created a tibble with all the probabilities from the distributions and plotted them using a bar plot
```{r}
#|size: scriptsize

# Binomial
binom.prob <- c(
  dbinom(0, size=2, prob=0.70), # Loses 0-2
  (2/3) * dbinom(1, size=3, prob=0.70), # Loses 1-2
  dbinom(2, size=2, prob=0.70), # Wins 2-0
  (2/3) * dbinom(2, size=3, prob=0.70) # Wins 2-1
)

# Negative Binomial
nbinom.prob <- c(
  dnbinom(0, size=2, prob=0.30), # Loses 0-2
  dnbinom(1, size=2, prob=0.30), # Loses 1-2
  dnbinom(0, size=2, prob=0.70), # Wins 2-0
  dnbinom(1, size=2, prob=0.70) # Wins 2-1
)

# Beta-Binomial
bbinom.prob <- c(
  dbbinom(0, size=2, alpha, beta), # Loses 0-2
  (2/3) * dbbinom(1, size=3, alpha, beta), # Loses 1-2
  dbbinom(2, size=2, alpha, beta), # Wins 2-0
  (2/3) * dbbinom(2, size=3, alpha, beta) # Wins 2-1
)

# Altham Binomial- theta = 0.50
alt05.prob <- c(
  daltbinom(0, size=2, p=0.70, t=0.5), # Loses 0-2
  (2/3) * daltbinom(1, size=3, p=0.70, t=0.5),# Loses 1-2
  daltbinom(2, size=2, p=0.70, t=0.5), # Wins 2-0
  (2/3) * daltbinom(2, size=3, p=0.70, t=0.5) # Wins 2-1
)

# Altham Binomial- theta = 1.50
alt15.prob <- c(
  daltbinom(0, size=2, p=0.70, t=1.5), # Loses 0-2
  (2/3) * daltbinom(1, size=3, p=0.70, t=1.5), # Loses 1-2
  daltbinom(2, size=2, p=0.70, t=1.5), # Wins 2-0
  (2/3) * daltbinom(2, size=3, p=0.70, t=1.5) # Wins 2-1
)

# Altham Beta-Binomial- theta = 0.50
altbb05.prob <- c(
  daltbbinom(0, size=2, t=0.5, a=alpha, b=beta), # Loses 0-2
  (2/3) * daltbbinom(1, size=3, t=0.5, a=alpha, b=beta), # Loses 1-2
  daltbbinom(2, size=2, t=0.5, a=alpha, b=beta), # Wins 2-0
  (2/3) * daltbbinom(2, size=3, t=0.5, a=alpha, b=beta) # Wins 2-1
)

# Altham Beta-Binomial- theta = 1.50
altbb15.prob <- c(
  daltbbinom(0, size=2, t=1.5, a=alpha, b=beta), # Loses 0-2
  (2/3) * daltbbinom(1, size=3, t=1.5, a=alpha, b=beta), # Loses 1-2
  daltbbinom(2, size=2, t=1.5, a=alpha, b=beta), # Wins 2-0
  (2/3) * daltbbinom(2, size=3, t=1.5, a=alpha, b=beta) # Wins 2-1
)

# in one tibble
gg.dat <- tibble(
  Outcome = c("Loses 0-2", "Loses 1-2", "Wins 2-0", "Wins 2-1"),
  Binom = binom.prob,
  Negative.Binom = nbinom.prob,
  Beta.Binom = bbinom.prob,
  Altham.Binom05 = alt05.prob,
  Altham.Binom15 = alt15.prob,
  Altham.BBinom05 = altbb05.prob,
  Altham.BBinom15 = altbb15.prob
)
(gg.dat)

# reshaping with pivot_long
gg.long <- gg.dat |>
  pivot_longer(
    cols = -Outcome,
    names_to = "Distribution",
    values_to = "Probability"
  )

# bar plot for all distribs
ggplot(gg.long, aes(x = Outcome, y = Probability)) +
  geom_col() +
  facet_wrap(~ Distribution) +
  theme_bw() +
  ylab("Probability") +
  xlab("Outcome")

```

**Comment**

The Binomial and Negative Binomial models give almost the same probabilities for the outcomes, with Player A more likely to win the match since they have a higher chance of winning each leg. The Beta-Binomial spreads the probabilities out a bit due to uncertainty in Player A’s skill, while the Altham models change how balanced the match looks depending on the value of theta.